mathml: Translate R expressions to MathML and LaTeX/MathJax

### 2022-12-18 Draft3.Rmd
support for multiple indices

The present R package defines a set of rules for an _automatic_ translation of
R expressions to mathematical output in RMarkdown documents [@Xie2020] and
ShinyApp webpages [@Chang2022]. The translation is done by an embedded Prolog
interpreter that maps nested expressions recursively to MathML and
LaTeX/MathJax, respectively. User-defined hooks enable to extend the set of
rules, for example, to represent specific R elements by custom mathematical signs. Moreover, in addition to graphical rendering, the same R expression provides its calculation or evaluation. This feature of the package saves the user's time and allows faster detection of errors.

@@

As is seen from the output above, the variable `term` is not assigned the result
of the calculation, but an R\ "call"
[see, e.g., @Wickham2019, for more details on "non-standard evaluation"]. This
call can eventually be evaluated with `eval(...)`,

@@

\pkg{mathml} contains several basic elements used in everyday mathematical
expressions, such as superscripts, subscripts, hat, relation and set signs,
accents and arrows. Below we provide some examples of their application with
commented code. The other basic elements will be displayed in
Table \@ref(tab:basic-elements) and Table \@ref(tab:arrows).

Let us start with the function for the linear regression. As the reader may notice, \pkg{mathml} displays Greek letters, accents and superscripts.

```{r, results='asis'}
term <- quote(hat(y) == hat(b)["0"] + hat(b)["1"]*X["1"])
mathout(term)

# or with greek letters as:
term <- quote(y== beta["0"] + beta["1"]*x["1"] + beta["2"]*x["2"] + ... + beta["i"]*x["i"] + varepsilon["i"])

mathout(term)
```

```{r basic-elements, echo=FALSE}
op <- list(subsup=quote(x["a"]^b), lchoose=quote(lchoose(n, k)), sqrt=quote(sqrt(x)), not=quote(!x), different=quote(x != y), mean=quote(mean(xy)), absolute=quote(abs(x)), tilde=quote(X ~ Y), approx=quote(x %~~% y), equiv=quote(x %==% y), congr=quote(x %=~% y), prop=quote(x %prop% y), 'in'=quote(a %in% A), intersect=quote(intersect(A, B)), union=quote(union(A, B)), intersect=quote(is.null(intersect(A, B))))
m <- lapply(op, FUN=mathjax, flags=list(cat=FALSE))

op2 <- list(twosidearrow=quote(x%<->%b), rightarrow=quote(x%->%y), leftarrow=quote(x%<-%y), uparrow=quote(x%up%y), down=quote(x%down%y), iff=quote(x%<=>%y), rArr=quote(x %=>% y), lArr=quote(x %<=% y), dblup=quote(x %dblup% y), dbldown=quote(x %dbldown% y))
m2 <- lapply(op2, FUN=mathjax, flags=list(cat=FALSE))

t <- cbind(Operator=op, Output=m, Operator=op2, Output=m2)
knitr::kable(t, caption="Basic elements in mathml", row.names=FALSE, escape=FALSE)
```

## Basic calculations

@@

A few more custom operators are shown in Table \@ref(tab:custom-operators).

```{r custom-operators, echo=FALSE}
op <- list(cross=quote(x %*% y), dot=quote(x %.% y), kronecker=quote(X %x% Y),
  'integer division'=quote(x %/% y), and=quote(A & B), vel=quote(A | B), xor=quote(xor(A, B)), not=quote(!A))
m <- lapply(op, FUN=mathjax, flags=list(cat=FALSE))

@@

>>>>

The present R package defines a set of rules for an _automatic_ translation of
R expressions to mathematical output in RMarkdown documents [@Xie2020] and
ShinyApp webpages [@Chang2022]. The translation is done by an embedded Prolog
interpreter that maps nested expressions recursively to MathML and
LaTeX/MathJax, respectively. User-defined hooks enable to extend the set of
rules, for example, to represent specific R elements by custom mathematical
signs.

The main feature of the package that the same R expressions and equations are
used for both mathematical typesetting and calculations. This saves time and
reduces mistakes, as well be illustrated below.

@@

As is seen from the output above, the variable `term` is not assigned the result
of the calculation, but an R\ "call"
[see, e.g., @Wickham2019, for details on "non-standard evaluation"]. This call
can eventually be evaluated with `eval(...)`,

@@

\pkg{mathml} contains several basic elements used in everyday mathematical
expressions, such as numbers, Latin and Greek letters, multi-letter identifiers,
accents, subscripts, and superscripts.

```{r, results='asis'}
term <- quote(1 + 2L + "a" + a + abc + phi + Phi + varphi + hat(b)[i, j]^2L
              + NaN + NA + TRUE + FALSE + Inf + -Inf)
mathout(term)
```

todo: parentheses around -infty, see also $(no)^{p+q}$ below.

## Operators and parentheses

@@

More operators are shown in Table \@ref(tab:custom-operators).

```{r custom-operators, echo=FALSE}
op <- list(
  cross=quote(x %*% y),
  dot=quote(x %.% y),
  kronecker=quote(X %x% Y),
  'integer division'=quote(x %/% y),
  and=quote(A & B),
  vel=quote(A | B),
  xor=quote(xor(A, B)),
  not=quote(!A),
  equal=quote(X == Y),
  assign=quote(X <- Y),
  different=quote(x != y),
  tilde=quote(X ~ Y),
  approx=quote(x %~~% y),
  equiv=quote(x %==% y),
  congr=quote(x %=~% y),
  prop=quote(x %prop% y),
  'in'=quote(a %in% A),
  intersect=quote(intersect(A, B)),
  union=quote(union(A, B)),
  intersect=quote(is.null(intersect(A, B)))) # todo: check if duplicate

m <- lapply(op, FUN=mathjax, flags=list(cat=FALSE))

@@

## R functions

A few more elements are displayed in
Tables \@ref(tab:basic-elements) and \@ref(tab:arrows).

```{r basic-elements, echo=FALSE}
op1 <- list(
  lchoose=quote(lchoose(n, k)),
  sqrt=quote(sqrt(x)),
  mean=quote(mean(xy)),
  absolute=quote(abs(x))) # todo: check if duplicate
m1 <- lapply(op1, FUN=mathout, flags=list(cat=FALSE))

op1 <- as.character(op1)
if(knitr::is_latex_output())
  op1 <- sapply(op1, FUN=knitr:::escape_latex)
if(knitr::is_html_output())
  op1 <- sapply(op1, FUN=knitr:::escape_html)

op2 <- list(
  twosidearrow=quote(x%<->%b),
  rightarrow=quote(x%->%y),
  leftarrow=quote(x%<-%y),
  uparrow=quote(x%up%y),
  down=quote(x%down%y),
  iff=quote(x%<=>%y),
  rArr=quote(x %=>% y),
  lArr=quote(x %<=% y),
  dblup=quote(x %dblup% y),
  dbldown=quote(x %dbldown% y))
m2 <- lapply(op2, FUN=mathout, flags=list(cat=FALSE))

t <- cbind(Operator=op1, Output=m1, Operator=op2, Output=m2)
knitr::kable(t, caption="Basic elements in mathml", row.names=FALSE, escape=FALSE)
```
